Estimating duration in the presence of misclassification

### 2024-03-25 introduction.tex
Updates from Overleaf

COVID-19 (coronavirus disease 2019), the disease caused by the virus SARS-CoV-2 (severe acute respiratory syndrome coronavirus 2), is estimated to have killed approximately 15 million in the most acute phase (to the end of 2021) and 30 million in total~\autocite{whoCOVIDExcess,economistCOVIDExcess}.
History suggests severe pandemics are not rare (see \cref{intro:fig:pandemic-timeline}).
\begin{figure}
    \includegraphics[width=\textwidth]{introduction/pandemic-timeline}
    \caption[Timeline of pandemics.]{%
        Timeline of pandemics.
        See \url{https://ourworldindata.org/historical-pandemics} for details and sources.
        Licensed under \href{https://creativecommons.org/licenses/by/4.0/}{CC-BY} by Saloni Dattani, Klara Auerbach, Marwa Boukarim, and Max Roser.
    }
    \label{intro:fig:pandemic-timeline}
\end{figure}
Fully quantifying the burden of historical pandemics is challenging due to the absence of systematic records.
However, our best records suggests that, since 1600, they have (on average) occurred every four years with a mean death toll of 5.6 million~\autocite{maraniNovelEpidemics}.
Recent estimates suggest that the current mean impact of pandemics is over 2.5 million deaths per year~\autocite{madhavPandemicMortality}.

@@

Reproduction numbers are strongly connected to \emph{growth rates}.
Over short periods of time, epidemics tend to grow (or shrink) exponentially, and hence have an exponential growth rate, normally denoted $r$.
For a given disease and population, an analogous growth rate can be defined for each definition of the reproduction number~\autocite{pellisEstimation,paragGrowthRates,wallingaGI}.
The advantage of using the growth rate it is easier to estimate and captures how quickly spread happens in calendar time; however, they do not directly give the strength of intervention required to control the epidemic~\autocite{royalSocietyRnumber}.

@@

\emph{Prevalence} is the proportion of the population with the disease at a given time.
Whether a given individual is \emph{prevalent} (has the disease) is often measurable through a diagnostic test.
Therefore, the prevalence of the population can be estimated with a prevalence survey, although these are typically very expensive to power appropriately.
Prevalence and incidence are strongly related, via the duration of an infection (see \cref{E-inc-prev}).

@@

Collecting data which directly informs inference of the quantities in this section is expensive or impossible.
Hence, they are typically inferred from other data sources through modelling.

@@

The data sources used for outbreak surveillance are often observational in nature.
Therefore, they suffer from biases that can be hard to understand and quantify~\autocite{shadboltChallenges}.
Here I discuss three types of data which are useful for surveillance.
Incidence-type and prevalence-type data can directly inform estimation of incidence and transmission; however, this estimation almost always relies on prior knowledge of epidemiological and/or biological quantities.
For example, the time between infection and symptom onset.
These quantities can be estimated from other studies, \eg cohort studies.

@@

Incidence-type data originates from observations of the consequences of a disease, \eg the development of symptoms or admission to hospital.
This type of data typically suffers from reporting biases, particularly that the proportion of infections which are counted can vary over time or within different subgroups of the population~\autocites[chapter 9]{lashModern}{shadboltChallenges}.

@@

However, severe symptoms can be rare in some population subgroups (\eg young, otherwise healthy individuals infected with SARS-CoV-2~\autocite{wardSero,bhopalChildren}).
This means that the observations do not provide much information on incidence in those subgroups.
Furthermore, the delay between infection and the severe event occurring means that recent infections cannot be estimated~\autocite{swallow2022challenges}.

@@

Prevalence-type data originates from observations of the number or proportion of a population in a specific disease state at a specific time.
For example, the number of hospital beds occupied by individuals with the disease or the proportion of the population with a marker of prior infection.
The issues associated with incidence-type data often also apply to prevalence-type data.

@@

Next, are semi-mechanistic models.
These have components of the transmission process in them but do not fully represent the process.
The most common type of semi-mechanistic model is those based on a \emph{renewal equation}~\autocite{bhattSemimechanistic}.
The renewal equation~\autocite{fraserRenewal} for an infectious disease, in the most general form, is:
\begin{align}
    Z(t) &= \int_{0}^{t} \beta(t,\tau) Z(t-\tau) d \tau
    \label{intro:eq:renewal}
\end{align}
where $Z(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the mean rate an individual infected $\tau$ time units ago infects individuals at time $t$.
Therefore, the renewal equation relates incidence in the past to incidence in the future.
This partially describes the transmission process, hence the name semi-mechanistic.
The effective reproduction number is $\Re = \int_0^\infty \beta(t, \tau) d\tau$.
% A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
% The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
Extensions of \cref{intro:eq:renewal} to track the number of susceptible individuals become full mechanistic models, equivalent to the compartmental models introduced next~\autocite{champredonEquivalence}.
Renewal equations do not feature further in this thesis; for details, including practical considerations when using them as the basis for inference, see \textcite{nashEstimating,gosticPractical}.

Finally, mechanistic models are the most complex, representing the transmission process in varying levels of detail.
\emph{Compartmental} models, the most common type of mechanistic model, split the population into compartments, each representing a different state of the disease.
The \emph{dynamics} of the epidemic, how it changes over time, are represented by the transitions between compartments (for more details see \cref{E-SEIR:sec:mechanistic-models}).
Simple mechanistic models use a homogeneous population, with individuals differing only due to their disease state, and exponential distributions for the time in each disease state.
More complex models relax these assumptions, in particular to incorporate heterogeneity.
The age structure of the population is important for respiratory diseases because individuals meet individuals of their own age more frequently.
% Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
% Unlike renewal equations, there are not widely-used implementations of compartmental models.
% A commonly used frameork is the SEIR (susceptible, exposed, infectious, recovered) model.
% The SEIR model, has four components (susceptible, exposed, infectious, recovered), with the population divided into these components; it is described in detail in \cref{E-SEIR:sec:mechanistic-models}.

@@

Solving \cref{intro:eq:disease-model} for $Z$ is an ill-posed problem when $f(0)$ is small.
The forwards error of the solution is large when the ratio of the off-diagonal to diagonal elements is large~\autocite[chapter 8.2]{highamAccuracy}; for this situation, which is when the value of $\max_{i > 0} f(i) / f(0)$ is large.
A large forwards error means that the computed value is far from the true solution~\autocite[6--9]{highamAccuracy}; this is avoided by either using an approach which gives a large $f(0)$ or smoothing $Z_t$.

@@

An observation model links $W_t$ to a possibly biased and/or noisy observation of it at time $t$.

@@

The biomarker approach models the trajectory of a biomarker over time in a cohort of individuals.
An individual is in the disease state while the biomarker is above or below some threshold.
This approach often requires parametric assumption(s).
For example, the functional form of the biomarker's trajectory.
In addition, if individual-level variation wants to be included, which is generally important, a distributional form of random effects is required.
This requires both sufficient observations per individual and an assumption on the form of the distribution.
More details are in \cref{E-ATACCC}.

@@

The surveillance systems used in response to SARS-CoV-2 are historically unique.
Modern technology allowed large amounts of data to be collected, distributed, and analysed.
While the HIV/AIDS pandemic inspired the development of a large amount of statistical methodology, the pandemic was far slower moving.
Much analysis of SARS-CoV-2 data was done at great haste, requiring trade-offs to be made between timeliness and rigour of results presented.

@@

For large parts of the pandemic, community testing was widely available in the UK.
However, it did not test a random or even representative sample of the population.
The number of tests conducted varied over time and space, for a variety of known and unknown reasons.
These include: policy on who is eligible for testing, the availability of tests, and the willingness of individuals to be tested.
Therefore, fluctuations in the number of positive tests may not reflect changes in the true incidence of the disease.
The large number of tests conducted allowed detailed analyses to be performed, including stratification by small geographic areas or age groups, although this does not solve the issue of the biases present.

Severe events counts are generally a reliable measure of the burden of the disease.
The criteria for being a severe event are normally well-defined and generally well-recorded in the UK context.
Therefore, these data are subject to fewer biases than community testing.
However, SARS-CoV-2 rarely leads to severe events in younger individuals, meaning severe event counts are insensitive to changes in transmission only occurring in younger individuals.

@@

The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys in representative samples of the general population.
These were not subject to the biases of community testing and observed the whole population, unlike severe events, although suffered from non-response bias.
Therefore, some modelling is required to borrow information across strata or time to provide stable estimates, as well as to correct for non-response bias (see \cref{E-intro:sec:cis}).
Despite this issue, they are considered the gold-standard metric.

@@

A second prevalence survey ran in the UK, known as REACT (REal-time Assessment of Community Transmission)~\autocite{rileyResurgence,rileyREACT}.
REACT was a series of cross-sectional surveys, run across 19 rounds between May 2020 and March 2022.
Each round consisted of at least \numprint{100000} randomly sampled individuals from the UK contributing swabs for testing and questionnaire data~\autocite{elliottTwin}.

@@

\Textcite{ealesCharacterising} estimated the duration of positivity from REACT data.
Specifically, individuals who tested positive in round 8 of REACT (6th--22 Jan 2021) were invited to take two additional swab tests.
Defining the time of the initial positive test at time 0, \textcite{ealesCharacterising} estimated the probability of remaining positive at each day after the initial positive test.
They made a strong parametric assumption that the probability decreases exponentially, with a sensitivity analysis assuming a plateau before the decay.
This assumption is necessary because they had at most 17 days of follow-up, and hence could not estimate the tail of the distribution.
Overall, they estimate a median duration of 14.0 days (95\% CI: 12.9--15.4 days) but with substantial positive correlation with each individual’s viral load at the time of the initial positive test (\ie a higher initial viral load was associated with a longer duration of positivity)
The individuals with the highest viral load individuals were estimated to have a median duration of 19.6 days (16.1--25.8 days).
The association could be due to individuals with a higher initial viral load being (on average) more recently infected at the time of the initial positive test.
The study design prohibits estimating how long before the initial positive test the infection occurred.
The same issue means the study underestimates the duration of positivity.

@@

In order to achieve these aims, I use a case study, specifically the SARS-CoV-2 epidemic in England from Mon 31 Aug 2020 until Sun 25 Jan 2021.
This case study has several features that make it useful for the aims of this thesis.
These include that the CIS was of sufficient size and incidence sufficiently high that there is enough data to perform the analyses.
Additionally, incidence begins rising over this period, which is the period when a surveillance system needs to provide a warning of upcoming issues.
The CIS began in April 2020, when incidence was falling due to the first lockdown in the UK.
Incidence remained low and fairly flat in England over summer 2020, until it began rising in mid or late August.
Finally, the period ends before the roll-out of the vaccine, which likely affects many important aspects of the analysis.
In particular, the duration of positivity and the susceptibility of the population.
Including both of these elements would require more complex modelling but include idiosyncratic features of the SARS-CoV-2 pandemic that are probably less generalisable to future diseases.

>>>>

COVID-19 (coronavirus disease 2019), the disease caused by the virus SARS-CoV-2 (severe acute respiratory syndrome coronavirus 2), is estimated to have killed approximately 15 million in the most acute phase (to the end of 2021) and 30 million in total~\autocite{whoCOVIDExcess,economistCOVIDExcess}.

History suggests severe pandemics are not rare (see \cref{intro:fig:pandemic-timeline}).
\begin{figure}
    \includegraphics[width=\textwidth]{introduction/pandemic-timeline}
    \caption[Timeline of pandemics.]{%
        Timeline of pandemics.
        See \url{https://ourworldindata.org/historical-pandemics} for details and sources.
        Licensed under \href{https://creativecommons.org/licenses/by/4.0/}{CC-BY} by Saloni Dattani, Klara Auerbach, Marwa Boukarim, and Max Roser.
    }
    \label{intro:fig:pandemic-timeline}
\end{figure}
Fully quantifying the burden of historical pandemics is challenging due to the absence of systematic records.
However, our best records suggests that, since 1600, they have (on average) occurred every four years with a mean death toll of 5.6 million~\autocite{maraniNovelEpidemics}.
Recent estimates suggest that the current mean impact of pandemics is over 2.5 million deaths per year~\autocite{madhavPandemicMortality}.

@@

Reproduction numbers are strongly connected to \emph{growth rates}.
Over short periods of time, epidemics tend to grow (or shrink) exponentially, and hence have an exponential growth rate, normally denoted $r$.
For a given disease and population, an analogous growth rate can be defined for each definition of the reproduction number~\autocite{pellisEstimation,paragGrowthRates,wallingaGI}.
The advantages of using the growth rates are that they are easier to estimate and capture how quickly spread happens in calendar time; however, they do not directly give the strength of intervention required to control the epidemic~\autocite{royalSocietyRnumber}.

@@

\emph{Prevalence} is the proportion of the population with the disease at a given time.
Whether a given individual is \emph{prevalent} (has the disease) is often measurable through diagnostic testing.
Therefore, the prevalence of the population can be estimated with a prevalence survey, although these are typically very expensive to power appropriately.
Prevalence and incidence are strongly related, via the duration of an infection (see \cref{E-inc-prev}).

@@

Collecting data that directly inform inference of the quantities in this section is expensive or impossible.
Hence, they are typically inferred from other data sources through modelling.

@@

Data sources used for outbreak surveillance are often observational in nature.
Therefore, they suffer from biases that can be hard to understand and quantify~\autocite{shadboltChallenges}.
Here I discuss three types of data which are useful for surveillance.
Incidence-type and prevalence-type data can inform estimation of incidence and transmission; however, this estimation almost always relies on prior knowledge of epidemiological and/or biological quantities.
For example, the time between infection and symptom onset.
These quantities can be estimated from other studies, \eg cohort studies.

@@

Incidence-type data originate from observations of the consequences of a disease, \eg the development of symptoms or admission to hospital.
This type of data typically suffers from reporting biases, particularly that the proportion of infections which are counted can vary over time or within different subgroups of the population~\autocites[chapter 9]{lashModern}{shadboltChallenges}.

@@

However, severe symptoms can be rare in some population subgroups (\eg young, otherwise healthy individuals infected with SARS-CoV-2~\autocite{wardSero,bhopalChildren}).
This means that the observations do not provide much information on incidence in those subgroups.
Furthermore, the delay between infection and the severe event occurring means that recent infections cannot be estimated with much precision~\autocite{swallow2022challenges}.

@@

Prevalence-type data originate from observations of the number or proportion of a population in a specific disease state at a specific time.
For example, the number of hospital beds occupied by individuals with the disease or the proportion of the population with a marker of prior infection.
The issues associated with incidence-type data often also apply to prevalence-type data.

@@

Next, are semi-mechanistic models.
These have components of the transmission process in them but do not fully represent the process.
A common type of semi-mechanistic model is those based on a \emph{renewal equation}~\autocite{bhattSemimechanistic}.
The renewal equation~\autocite{fraserRenewal} for an infectious disease, in the most general form, is:
\begin{align}
    Z(t) &= \int_{0}^{t} \beta(t,\tau) Z(t-\tau) d \tau
    \label{intro:eq:renewal}
\end{align}
where $Z(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the mean rate at which an individual infected $\tau$ time units ago infects individuals at time $t$.
Therefore, the renewal equation relates incidence in the past to incidence in the future.
This partially describes the transmission process, hence the name semi-mechanistic.
The effective reproduction number is $\Re = \int_0^\infty \beta(t, \tau) d\tau$.
% A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
% The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
Extensions of \cref{intro:eq:renewal} that track the number of susceptible individuals become full mechanistic models, equivalent to the compartmental models introduced next~\autocite{champredonEquivalence}.
Renewal equations do not feature further in this thesis; for details, including practical considerations when using them as the basis for inference, see \textcite{nashEstimating,gosticPractical}.

Finally, mechanistic models are the most complex models, representing the transmission process in varying levels of detail.
\emph{Compartmental} models, the most common type of mechanistic model, split the population into compartments, each representing a different state of the disease.
The \emph{dynamics} of the epidemic, how it changes over time, are represented by the transitions between compartments (for more details see \cref{E-SEIR:sec:mechanistic-models}).
Simple mechanistic models use a homogeneous population, with individuals differing only due to their disease state, and exponential distributions for the time in each disease state.
More complex models relax these assumptions, in particular to incorporate heterogeneity.
The age structure of the population is important for respiratory diseases because individuals meet individuals of their own age more frequently.
% Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
% Unlike renewal equations, there are not widely-used implementations of compartmental models.
% A commonly used frameork is the SEIR (susceptible, exposed, infectious, recovered) model.
% The SEIR model, has four components (susceptible, exposed, infectious, recovered), with the population divided into these components; it is described in detail in \cref{E-SEIR:sec:mechanistic-models}.

@@

Solving \cref{intro:eq:disease-model} for $Z$ is an ill-posed problem when $f(0)$ is small.
The forwards error of the solution is large when $\max_{i > 0} f(i) / f(0)$~\autocite[chapter 8.2]{highamAccuracy}.
A large forwards error means that the computed value is far from the true solution~\autocite[6--9]{highamAccuracy}; this is avoided by either using an approach which gives a large $f(0)$ or smoothing $Z_t$.

@@

Observation models link $W_t$ to a possibly biased and/or noisy observation of it at time $t$.

@@

The biomarker approach models the trajectory of a biomarker over time in a cohort of individuals.
An individual is in the disease state while the biomarker is above or below some threshold.
This approach often requires parametric assumption(s) such as a functional form of the biomarker's trajectory.
In addition, if individual-level variation is included, which is generally important, a distributional form of random effects is required.
This requires both sufficient observations of the biomarker per individual and an assumption on the form of the distribution.
More details are in \cref{E-ATACCC}.

@@

The surveillance systems used in response to SARS-CoV-2 are historically unique.
Modern technology allowed large amounts of data to be collected, distributed, and analysed.
While the HIV/AIDS pandemic inspired the development of a large amount of statistical methodology, it was far slower moving.
Much analysis of SARS-CoV-2 data was done at great haste, requiring trade-offs to be made between timeliness and rigour of results presented.

@@

For large parts of the pandemic, community testing was widely available in the UK.
However, it did not test a representative sample of the population.
The number of tests conducted varied over time and space, for a variety of known and unknown reasons.
These include: policy on who is eligible for testing, the availability of tests, and the willingness of individuals to be tested.
Therefore, fluctuations in the number of positive tests may not reflect changes in the true incidence of the disease.
The large number of tests conducted allowed detailed analyses to be performed, including stratification by small geographic areas or age groups, although this does not solve the issue of the biases present.

Severe events counts are generally a more reliable measure of the burden of the disease.
The criteria for being a severe event are normally well-defined and generally well-recorded in the UK context.
Therefore, these data are subject to fewer biases than community testing.
However, SARS-CoV-2 rarely leads to severe events in younger individuals, meaning severe event counts are insensitive to changes in transmission only occurring in younger individuals.

@@

The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys in representative samples of the general population.
These were designed to avoid the biases of community testing and observe the whole population.
Therefore, some modelling is required to borrow information across strata or time to provide stable estimates, as well as to correct for non-response bias (see \cref{E-intro:sec:cis}).
Despite this issue, they are considered the gold-standard metric.

@@

A second prevalence survey ran in the UK, known as REACT (REal-time Assessment of Community Transmission)~\autocite{rileyResurgence,rileyREACT}.
REACT was a series of cross-sectional surveys, run across 19 rounds between May 2020 and March 2022.
Each round consisted of at least 100,000 randomly sampled individuals from the UK contributing swabs for testing and questionnaire data~\autocite{elliottTwin}.

@@

\Textcite{ealesCharacterising} estimated the duration of positivity from REACT data.
Specifically, individuals who tested positive in round 8 of REACT (6th--22nd Jan 2021) were invited to take two additional swab tests.
Defining the time of the initial positive test at time 0, \textcite{ealesCharacterising} estimated the probability of remaining positive at each day after the initial positive test.
They made a strong parametric assumption that the probability decreases exponentially, with a sensitivity analysis assuming a plateau before the decay.
This assumption is necessary because they had at most 17 days of follow-up, and hence could not estimate the tail of the distribution.
Overall, they estimate a median duration of 14.0 days (95\% CI: 12.9--15.4 days) but with substantial positive correlation with each individual’s viral load at the time of the initial positive test (\ie a higher initial viral load was associated with a longer duration of positivity)
The individuals with the highest viral load individuals were estimated to have a median duration of 19.6 days (16.1--25.8 days).
The association could be due to individuals with a higher initial viral load being (on average) more recently infected at the time of the initial positive test.
The study design prohibits estimating how long before the initial positive test the infection occurred.
The same issue means the study underestimates the duration of positivity.

@@

In order to achieve these aims, I use a case study, specifically the SARS-CoV-2 epidemic in England from Mon 31st Aug 2020 until Sun 25th Jan 2021.
This case study has several features that make it useful for the aims of this thesis.
These include that the CIS was of sufficient size and incidence sufficiently high that there is enough data to perform the analyses.
Additionally, incidence begins rising over this period, which is the period when a surveillance system needs to provide a warning of upcoming issues.
The CIS began in April 2020, when incidence was falling due to the first lockdown in the UK.
Incidence remained low and fairly flat in England over summer 2020, until it began rising in mid or late August.
Finally, the period ends before the roll-out of the vaccine, which likely affects many important aspects of the analysis.
In particular, the duration of positivity and the susceptibility of the population.
Including both of these elements would require more complex modelling but include idiosyncratic features of the SARS-CoV-2 pandemic that are probably less generalisable to future diseases.
